---
title: "Compare All .csv files"
author: "mcc"
date: "6/14/2020"
output: html_document
---

### This is Rmarkdown file shows the preliminary Exploratory Data Analysis for

| No. | File Name     | Abbreviation |
| :-- | :------------ | :----------- |
| 1   | AFP11.csv     | AFP          |
| 2   | Non-AFP11.csv | Non_AFP      |

I will not use the two alternative files because their is no information that I can even glean off the filenames.

Data found at https://www.kaggle.com/rafay12/anti-freeze-protein-classification/metadata.

### Import libraries
```{r message=FALSE, warning=FALSE}
library(easypackages)
libraries("knitr", "readr", "caret", "doMC")

opts_chunk$set(echo = TRUE)
```

### Load data
```{r message=FALSE, warning=FALSE}
AFP <- read_csv("AFP11.csv", 
                col_names = FALSE,
                col_types = cols(X2 = col_factor(levels = c("0","1"))))
                
Non_AFP <- read_csv("Non-AFP11.csv"
                    col_names = FALSE,
                    col_types = cols(X2 = col_factor(levels = c("0","1"))))
```

### AFP11 & Non_AFP11 Dimensions
```{r}
dim(AFP)
```

```{r}
dim(Non_AFP)
```

### Check AFP11 & Non_AFP11 for missing values
```{r}
apply(is.na(AFP), 2, which)
```

```{r}
apply(is.na(Non_AFP), 2, which)
```

Note: NO missing values indicated.

### Boxplots of Columns x3-x22 of AFP11 & Non_AFP11 datasets
```{r}
# Subset out columns 2 through 22 for first summary
afp_3_22 <- AFP[c(3:22)] 
non_afp_3_22 <- Non_AFP[c(3:22)] 
```

Boxplots: Columns x2-x22 of AFP11 dataset
```{r}
boxplot(afp_3_22, 
        main = "Boxplots: AFP Columns x3-x22 - (%AAC?)",
        ylab = "% AAC(?)",
        xlab = "Amino Acid",
        las = 1)
```
Boxplots: Columns x2-x22 of AFP11 dataset
```{r}
boxplot(non_afp_3_22, 
        main = "Boxplots: NON-AFP Columns x3-x22 - (%AAC?)",
        ylab = "% AAC(?)",
        xlab = "Amino Acid",
        las = 1)
```

Inspection of Columns, x3 to x842

Note 5: Columns x23 to x423 may be dipeptides, since 20 x 20 = 400. 

Other common measures may include but not limited to:

| No. | Name | Reference |
| :-- | :--- | :-------- |
| 1 | Pseudo amino acid composition, or PseAAC | Reference [^1] |
| 2 | Hydrophobicity | Reference |
| 3 | Hydrophilicity, or PseAAC | Reference |
| 4 | pK1 (alpha-COOH) | Reference |
| 5 | pK2 (NH3) | Reference |
| 6 | pI (at 25C) | Reference |
| 7 | Autocorrelation Normalized Moreau-Broto Autocorrelation | Reference [^] |
| 8 | Moran Autocorrelation | Reference [^] |
| 9 | Geary Autocorrelation | Reference [^] |
| 10 | CTD Composition | Reference [^] |
| 11 | Transition |  Reference [^] |
| 12 | Distribution | Reference [^] |
| 13 | Conjoint Triad | Reference [^] |
| 14 | Quasi-Sequence-Order | Reference [^] |
| 15 | Sequence-Order-Coupling Number | Reference [^] |
| 16 | Quasi-Sequence-Order Descriptors | Reference [^] |
| 17 | Amphiphilic Pseudo-Amino Acid Composition | Reference [^] |
| 18 | Amino acid composition, AAC | Reference [^] |

See also: ProtrWeb: http://protr.org/

[^1]:Hong-Bin Shen and Kuo-Chen Chou. PseAAC: a flexible web-server for generating various kinds of protein pseudo amino acid composition. Analytical Biochemistry, 2008, 373: 386-388


## Check for Zero-Variance in Column Features
```{r}
col_variances <- apply(AFP[c(3:1742)], 2, var)
'0' %in% col_variances

# which(col_variances %in% 0) 

summary(col_variances)
```

```{r}
col_variances <- apply(Non_AFP[c(3:1742)], 2, var)
#'0' %in% col_variances

# which(col_variances %in% 0) 

summary(col_variances)
```

| Dataset |  Min Var. |  Max Var. | Difference of Log10 |
|--------:|----------:|----------:|--------------------:|
| AFP     |  0.000006 | 15.735878 |              3.1976 |
| Non_AFP | 0.0009145 | 1.4412858 |              6.4187 |

Therefore delete the 2nd column of the original, AFP11, dataframe.

### Zero- and Near Zero-Variance Predictors Using **caret**
```{r}
nzv <- nearZeroVar(afp_3_22, saveMetrics = TRUE)
head(nzv)
```



### Identifying Correlated Predictors
```{r}
correlation_matrix <-  cor(afp_3_22)

summary(correlation_matrix[upper.tri(correlation_matrix)])

# If the summary of the correlation matrix had max or min values greater than 0.75 and -0.75 the we could find the locations of the pairs with relative high correlationos.
#high_correlation <- sum(abs(correlation_matrix[upper.tri(correlation_matrix)]) > 0.75)
```






