---
title: "Anti-freeze-protein-classification EDA & PCA"
author: 'Matthew Curcio'
date: '10/8/2020'
output:
  html_document:
    number_sections: true
    toc: true
    fig_width: 7
    fig_height: 4.5
    theme: readable
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

# Abstract

# Introduction

I have chosen to work with R and the Anti-freeze-protein-classification dataset. [^18]

I will focus on using Zero Variance and Near-Zero Variance as a tool for reducing dimensions.
Then move on to Principal Component Analysis where the optimum number of principal components will be determined using XGBoost as a measure. 

I am new to machine learning so feedback is welcome! 

There are 5 parts to my report:

1. Exploratory Data Analysis
1. Near-Zero Variance Cut-off
1. Principal Component Analysis
1. Conclusion

[^18]: https://www.kaggle.com/rafay12/anti-freeze-protein-classification



The files listed below are presumed to be a positive and negative control for anti-freeze proteins.

| File Name     | Working Abbreviation | Observations | Columns (Features) | Missing Values | File Size |
| :------------ | :------------------  | -----------: | -----------------: | -------------: | --------: |
| AFP11.csv     |                  afp |          481 |         1742(1740) |              0 |    4.8 MB |
| Non-AFP11.csv |              non_afp |         9191 |         1742(1740) |              0 |   91.6 MB |


Column Variables appear to be set up as:

| Column(s) | Structure | Notes / Description                           |
| :-------: | :-------- | :-------------------------------------------- |
|        X1 | character | Protein ID; <br> Uniprot accession numbers(?) |
|        X2 | numerical | Column contains zeros. <br> It would be useful to change this column to a **label variable set of [0,1]** |
|    X3:X23 | numerical | Amino acid percent composition(AAC), order unknown; <br> range = [0, 25.904] |
| X24:X1742 | numerical | Alternative Descriptors, order unknown; <br> range = [0, 50]; <br> See Table #1 |

Note 1: At this point, I am unsure how these files were derived. However these values look consistent with amino acid composition and Alternative Descriptors analysis.
    
Note 2: The data sets do not appear to have amino acid total count (protein length) data. I find this information useful when one encounters amino acid compositions greater than 10%. The protein length can be used as a feature but more often it is a good visual check to have.


## Table #1: Non-Exhuastive List of Protein Descriptors:

| No. | Descriptor Name                                      | Notes                          |
| :-: | :--------------------------------------------------- | :----------------------------- |
|  1  | Pseudo amino acid composition (PseAAC)               | K.C. Chou [^1]                 |
|  2  | Hydrophobicity                                       | C. Tanford [^2]                |
|  3  | Hydrophilicity                                       | T.P. Hopp [^4]                 |
|  4  | pK1 (alpha-COOH)                                     | CRC Handbook of Chemistry [^5] |
|  5  | pK2 (-NH3)                                           | R.M.C. Dawson [^6]             |
|  6  | pI (at 25C)                                          | R.M.C. Dawson [^6]             |
|  7  | Moreau-Broto Autocorrelation                         | G. Moreau, P. Broto [^7] [^19] |
|  8  | Moran Autocorrelation                                | P. Moran [^8] [^20]            |
|  9  | Geary Autocorrelation                                | R.C. Geary [^9]                |
| 10  | Composition, Transition and Distribution (CTD) of AA | G. Govindan [^10]              |
| 11  | Transition                                           | [^11]                          |
| 12  | Distribution                                         | [^12]                          |
| 13  | Conjoint Triad                                       | J.W. Shen [^13]                |
| 14  | Quasi-Sequence-Order                                 | K.C. Chou [^14]                |
| 15  | Sequence-Order-Coupling Number                       | [^15]                          |
| 16  | Quasi-Sequence-Order Descriptors                     | [^16]                          |
| 17  | Amphiphilic Pseudo-Amino Acid Composition            | [^17]                          |

See also: ProtrWeb: http://protr.org/   https://nanx.me/papers/protr.pdf

[^1]: Hong-Bin Shen, Kuo-Chen Chou. PseAAC: a flexible web-server for generating various kinds of protein pseudo amino acid composition. Analytical Biochemistry, 2008, 373: 386-388
[^2]: C. Tanford, JACS, 1962, 84: 4240-4246
[^4]: T.P.Hopp, K.R.Woods, PNAS, 1981, 78:3824-3828
[^5]: CRC Handbook of Chemistry and Physics, 66th ed., CRC Press, Boca Raton, Florida, 1985
[^6]: R.M.C. Dawson, D.C. Elliott, W.H. Elliott, K.M. Jones, Data for Biochemical Research 3rd ed., Clarendon Press Oxford (1986)
[^7]: http://www.rguha.net/writing/notes/desc/node2.html
[^19]: G. Moreau and P. Broto, Autocorrelation of a topological structure: A new molecular descriptor, Nouv. J. Chim. 4 (1980) 359–360
[^8]: Moran, P. A. P. (1950), "Notes on Continuous Stochastic Phenomena". Biometrika. 37 (1): 17–23
[^20]: https://pro.arcgis.com/en/pro-app/tool-reference/spatial-statistics/h-how-spatial-autocorrelation-moran-s-i-spatial-st.htm
[^9]: Geary, R. C. (1954). "The Contiguity Ratio and Statistical Mapping". The Incorporated Statistician. The Incorporated Statistician. 5 (3): 115–145
[^10]: Geetha Govindan; Achuthsankar S. Nair, Composition, Transition and Distribution (CTD) — A dynamic feature for predictions based on hierarchical structure of cellular sorting, IEEE, 16-18 Dec. 2011
[^11]: X
[^12]: X
[^13]: J.W. Shen, J. Zhang, X.M. Luo, W.L. Zhu, K.Q. Yu, K.X. Chen, Y.X. Li, H.L. Jiang. Predicting Protein-protein Interactions Based Only on Sequences Information. Proceedings of the National Academy of Sciences. 007, 104, 4337-4341
[^14]: Kuo-Chen Chou. Prediction of Protein Subcellar Locations by Incorporating Quasi-Sequence-Order Effect. Biochemical and Biophysical Research Communications, 2000, 278, 477-483
[^15]: A
[^16]: B
[^17]: C

**Using the Bash** command line I was able to quickly inspect the 2 datafiles. 

Note 3: The dataset does not appear to have column names as the first row.

# Exploratory Data Analysis

## Import libraries
```{r message=FALSE, warning=FALSE}
library(easypackages)
libraries("knitr", "readr", "caret", "kableExtra", "doMC",
          "ROCR" ,"caret" ,"ggplot2" ,"xgboost" ,"Metrics")
```

## Load data
```{r message=FALSE, warning=FALSE}
afp     <- read_csv("AFP11.csv", col_names = FALSE)
non_afp <- read_csv("Non-AFP11.csv", col_names = FALSE)
```

## Check no. dimensions
```{r}
dim(afp)

dim(non_afp)
```
Note 3: rows & columns

## Check for missing values
```{r}
sum(is.na(afp))

sum(is.na(non_afp))
```

Note 4: No missing values indicated.

## Set indicator variable (X2) to anti-freeze-proteins (1) & Non-anti-freeze-proteins (0)
```{r}
afp$X2     <- rep(1, 481)  # Set indicator variable to 1 for anti-freeze-proteins
non_afp$X2 <- rep(0, 9191) # Set indicator variable to 0 for NON-anti-freeze-proteins

all_proteins <- rbind(afp, non_afp)

dim(all_proteins)
```

## Amino Acid Composition: Numerical range of columns X3 - X23

```{r}
aac <- all_proteins[c(3:23)] 

ranges_aac <- range(aac)

ranges_aac
```

## Protein Descriptors: Numerical range of columns X24 - X1742

```{r}
dpc <- all_proteins[c(24:1742)] 

ranges_dpc <- range(dpc)

ranges_dpc
```

Note 5: It is recommended that all columnar variables (X3 - X1742) be normalized.

# Zero Variance & Near-Zero Variance Testing

## Check for Near-Zero Variance using Caret

```{r}
df <- all_proteins[c(3:1742)]
nzv <- nearZeroVar(df, saveMetrics = TRUE)
head(nzv)
```

## Table of Zero Variance Variables

```{r}
table(nzv$zeroVar)
```

Note 6: No Zero-Variance features. FALSE indicates all features do not have zero variance.

## Table of Near-Zero Variance Variables
```{r}
table(nzv$nzv)
```


## Summary of Near-Zero Variance to find 1st quartile cut-off
```{r}
summary(nzv$percentUnique)
```


```{r}
# How many features have Near-Zero Variance
print(length(nzv[nzv$zeroVar == T, ]))

print(paste('Column count before cutoff:', ncol(df)))

# How many have less than 4.446 `percentUnique`
dim(nzv[nzv$percentUnique > 4.446,])
```

## Remove zero and bottom quartile of near-zero variance
```{r}
# remove zero & near-zero variance from df_nzv
df_nzv <- df[c(rownames(nzv[nzv$percentUnique > 4.446, ])) ]

print(paste('Column count after cutoff:', ncol(df_nzv)))
```


# Principal Component Analysis

## Run `prcomp` on `df_nzv` (1302 features)

```{r}
registerDoMC(cores = 10) # Start multi-processor mode
start_time <- Sys.time() # Start timer

pmatrix <- scale(df_nzv)
pca_model <- prcomp(pmatrix)

registerDoSEQ() # Stop multi-processor mode
end_time <- Sys.time() # End timer
end_time - start_time # Display elapsed time
```


```{r}
plot(cumsum(pca_model$sdev^2 / sum(pca_model$sdev^2)), 
     type = "l",
     main = "PCA Model: Cumulative Proportion of Variances",
     xlab = "Feature No.",
     ylab = "Cumulative Proportion",
     xlim = c(0, 1350))
abline(v = 1160, col = "red")
```


See: https://github.com/amunategui/pca-dimension-reduction/blob/master/pca.R

```{r include=FALSE}
Evaluate_SE <- function(dfEvaluate) {
        CVs <- 5
        cvDivider <- floor(nrow(dfEvaluate) / (CVs+1))
        indexCount <- 1
        outcomeName <- c('cluster')
        predictors <- names(dfEvaluate)[!names(dfEvaluate) %in% outcomeName]
        lsErr <- c()
        lsAUC <- c()
        for (cv in seq(1:CVs)) {
                print(paste('cv', cv))
                dataTestIndex <- c((cv * cvDivider):(cv * cvDivider + cvDivider))
                dataTest  <- dfEvaluate[ dataTestIndex,]
                dataTrain <- dfEvaluate[-dataTestIndex,]
                
                bst <- xgboost(data = as.matrix(dataTrain[, predictors]),
                               label = dataTrain[, outcomeName],
                               max.depth = 6, eta = 1, verbose = 0,
                               nround = 5, nthread = 4, 
                               objective = "reg:squarederror")
                
                predictions <- predict(bst, 
                                       as.matrix(dataTest[, predictors]), 
                                       outputmargin = TRUE)
                err <- rmse(dataTest[,outcomeName], predictions)
                lsErr <- c(lsErr, err)
                gc()
        }
        print(paste('Mean Error:', mean(lsErr)))
}
```

## Standard 1: Run model on `df_nzv` data set

```{r echo=TRUE}
dfEvaluate <- cbind(as.data.frame(sapply(df_nzv, as.numeric)),
                    cluster = all_proteins$X2)

Evaluate_SE(dfEvaluate)
```



```{r eval=FALSE, include=FALSE}



nComp <- 5  
nComp <- 10  
nComp <- 90     
nComp <- 20  
nComp <- 50   
nComp <- 100   

# change nComp to try different numbers of component variables (10 works great)
nComp <- 10  # 0.9650
dfComponents <- predict(princ, 
                        newdata = pmatrix)[, 1:nComp]
dfEvaluate <- cbind(as.data.frame(dfComponents),
                    cluster = all_proteins$X2)

EvaluateAUC(dfEvaluate)
```





