---
title: "Compare All .csv files"
author: "mcc"
date: "6/14/2020"
output: html_document
---

### This is Rmarkdown file shows the preliminary Exploratory Data Analysis for

| No. | File Name     | Abbreviation |
| :-- | :------------ | :----------- |
| 1   | AFP11.csv     | AFP          |
| 2   | Non-AFP11.csv | Non_AFP      |

I will not use the two alternative files because their is no information that I can even glean off the filenames.

Data found at https://www.kaggle.com/rafay12/anti-freeze-protein-classification/metadata.

### Import libraries
```{r message=FALSE, warning=FALSE}
library(easypackages)
libraries("knitr", "readr", "caret", "doMC")
```

### Load data
```{r message=FALSE, warning=FALSE}
afp <- read_csv("AFP11.csv", col_names = FALSE)
non_afp <- read_csv("Non-AFP11.csv", col_names = FALSE)

all_afp <- rbind(afp, non_afp)
dim(all_afp)
```



### Zero- and Near Zero-Variance Predictors Using **caret**
```{r}
afp_scaled <- scale(df, center = TRUE, scale = TRUE)

nzv <- nearZeroVar(afp_3_22, saveMetrics = TRUE)
head(nzv)
```



### Identifying Correlated Predictors
```{r}
correlation_matrix <-  cor(afp_3_22)

summary(correlation_matrix[upper.tri(correlation_matrix)])

# If the summary of the correlation matrix had max or min values greater than 0.75 and -0.75 the we could find the locations of the pairs with relative high correlationos.
#high_correlation <- sum(abs(correlation_matrix[upper.tri(correlation_matrix)]) > 0.75)
```






